 

### 2021-05-24 vertical-thermal-review.Rmd
worked on biophysical section, removed sub-headings

### Foliage density
**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and in turn is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest.
Leaf area density.... *[description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references).]*
Vertical profiles in leaf area density vary across forests, being... *summarize some key differences across forest types* (REFS), and are altered (*HOW?*) following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (Condit, Meakem, ?any work on this in temperate forests??).
In this review, we focus on growing season conditions unless otherwise noted.

### Light environments

@@

### Wind speeds 
**Wind speeds are also higher at the top of the canopy, owing to... [@juckerCanopyStructureTopography2018].**
Across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites, wind speeds increase with height. 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

### Air temperature 
**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
The $T_{air}$ gradient along canopies and understories vary with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
Seasonally, canopy buffering capacity is greater in wet than dry seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al. 2017; **double check these refs--at least one didn't belong**].

### Humidity and evaporative demand
**Relative humidity, $RH$, is generally highest in the understory and decreases with height across the vertical profile [Fig. `r fig_NEON_vertical`; e.g., @mcgregor_tree_nodate].**
Along with temperature, dense-canopy forests maintain higher daily maximum $RH$ below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a]. 
$RH$, in combination with temperature, determines vapor pressure deficit [$VPD$] -- the driving force of water loss from a leaf, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Increased tree height [upto 20m with higher LAI] can also lower mean and max temperature and $VPD$ in the understory, beyond a certain height the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].

### Carbon dioxide

@@

### Geographic and climatic factors

**The buffering patterns described above vary geographically and across forests types.**
The strength of buffering decreases with altitude (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) and moving to open-forests [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], greater in wetter than drier regions [**REF**] and increases with proximity to water, for example in riparian forests. [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]

>>>>

**Canopy foliage acts as the primary physical barrier between the atmosphere and the forest floor, buffering multiple aspects of the understory conditions. It is critical in influencing -- and in turn is influenced by -- the vertical biophysical gradient (Fig. `r fig_NEON_vertical`).**
The distribution of this foliage is structured in correspondence with the distribution of tree sizes within a forest.
Leaf area density.... *[description of most common vertical pattern (@bonan_ecological_2016 reviews this and points to appropriate references).]*
Vertical profiles in leaf area density vary across forests, being... *summarize some key differences across forest types* (REFS), and are altered (*HOW?*) following disturbance [e.g., @parkerCanopyLightTransmittance2002 ; @almeidaContrastingFireDamage2016 ; @starkReframingTropicalSavannization2020].
They also vary seasonally, ... (Parker & Tibbs 2004; @smithSeasonalDroughtrelatedChanges2019 ; @parkerConsequencesEnvironmentalHeterogeneity2019). 
In mixed deciduous- evergreen broadleaf forests, the fraction of deciduous trees tends to increase across tree size classes (Condit, Meakem, ?any work on this in temperate forests??).
In this review, we focus on growing season conditions unless otherwise noted.

@@

**Wind speeds are also higher at the top of the canopy, owing to... [@juckerCanopyStructureTopography2018].**
Wind speed increase with height across forest types from closed-canopy forests [@barnardSeasonalVariationCanopy2016a; @mcgregor_tree_nodate] to open savannas [@curtis_intracanopy_2019], including all **#** forested NEON sites. 
Specifically, across the NEON sites, July maximum daily wind speeds averaged up to **~4** m s^-1^ higher at the top of the NEON vertical profile than at the bottom, where maximum wind speeds consistently averaged < **#** m s^-1^ (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_a$) and associated potential for loss of both water and heat (see next section).

**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
The $T_{air}$ gradient along canopies and understories vary with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. 
Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies -- *i.e.*, dense canopy understories can have cooler maximum daytime $T_{air}$ and warmer minimum $T_{air}$ during the night than open canopy understories or nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009]. 
(*Noteably, this pattern can be reversed in open forests, where...[explain])*

**Relative humidity, $RH$, is generally highest in the understory and decreases with height across the vertical profile [Fig. `r fig_NEON_vertical`; e.g., @mcgregor_tree_nodate].**
Dense-canopy forests maintain higher daily maximum $RH$ below-canopy than open forests and nearby open areas, and this effect is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a]. 
In combination with $T_{air}$, $RH$ determines vapor pressure deficit ($VPD$) -- the driving force of water loss from a leaf, which tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Increased tree height (up to 20 m with higher LAI) can also lower mean and maximum $T_{air}$ and $VPD$ in the understory, beyond a certain height the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. *( This last sentence needs work.)*

@@

**The buffering patterns described above vary geographically, seasonally, and across forests types [REFS; *Defrenne et al. 2021*].**
The strength of buffering decreases with altitude (**THIS WILL NOT ALWAYS BE TRUE!--e.g., semi-arid regions where forest cover increases with elevation**) and moving to open-forests [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], greater in wetter than drier regions [**REF**] and increases with proximity to water, for example in riparian forests. [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]
Seasonally, canopy buffering capacity is greater in wet than dry seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018, Tymen et al. 2017; **double check these refs--at least one didn't belong**]. *(Presumably, the above is because leaf area is greater in set seasons)*
